---
title: "Example Syllabus for 10 week course"
output: html_document
---


This is a syllabus for the 10 week course we taught at the University of Washington. Each week included two 80 minute periods. We rarely lectured for more than 15 at a time and incorporated many group activities. 

Students were required to have an introductory level statistics course, but we motivated and taught testing and estimation in a way that did not require previous experience with these concepts.

The course drew students from a range of disciplines including statistics, economics, fisheries, communications, social work,... While the bulk of the students were upper level undergraduates, we always had a few graduate students. This breadth of experience and expertise contributed to the course, and led to many intesting group discussions. 

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(tidyr)
library(dplyr)
library(kableExtra)
library(pander)
library(stringr)

s2020 <- read.csv("tables/SYTYCDS_syllabus_2020.csv",stringsAsFactors=FALSE)

s2020[,c("Week","Theme","Description")] %>% 
  mutate_all(function(x) stringr::str_replace_all(x, "\\n", "<br>"))%>%
  knitr::kable(escape=FALSE) %>%
  kable_styling(c("striped","bordered"), full_width = F, position="left")
```


```{r, echo=FALSE, results='asis'}
nList <- c("Tuesday","Thursday","Readings","Homework","Group")

res <- lapply(s2020$Week, function(x) {
  tmpText <- stringr::str_replace_all(t(s2020[x,nList]), "\\n", "; ")
  pDat <- data.frame(nList,tmpText)
  knitr::knit_child(text = c(
    '```{r echo=FALSE}',
    'knitr::kable(pDat,escape=FALSE,row.names=FALSE,col.names=c(paste("Week",x),"Details")) %>%  kable_styling(c("striped","bordered"), full_width = F, position="left")',
    '```',
    ''
  ), envir = environment(), quiet = TRUE)
})
cat(unlist(res), sep = '\n')
```

```{r eval=FALSE, echo=FALSE}
library(tidyr)
library(dplyr)
library(kableExtra)
addText <- function(x){
  tmp <- stringr::str_replace_all(x, "\\n", "\n|    ")
  tmp2 <- stringr::str_replace_all(tmp, ";", "\n|    ")
  ifelse(tmp2=="","",paste("|    ",tmp2,sep=""))
}

res <- lapply(s2020$Week, function(x) {
  knitr::knit_child(text = c(
    '**Week `r x`**\n',
    '|  **Day 1**',
    addText(s2020$Tuesday[x]),
    ''
  ), envir = environment(), quiet = TRUE)
})
cat(unlist(res), sep = '\n')


res <- lapply(s2020$Week, function(x) {
  pDat <- data.frame(nList,t(s2020[x,nList]))
  knitr::knit_child(text = c(
    '**Week `r x`**\n',
    '```{r echo=FALSE}',
    'knitr::kable(pDat,escape=FALSE,row.names=FALSE,col.names=c(paste("Week",x),"Details")) %>%  kable_styling(c("striped","bordered"), full_width = F)',
    '```',
    ''
  ), envir = environment(), quiet = TRUE)
})
cat(unlist(res), sep = '\n')

```


```{r eval=FALSE, echo=FALSE}
template <- "### Week %i
| Day 1 
|  
Additional section text is: %s.

" # dont't forget the newline

for (i in seq(nrow(input))) {
  current <- input[i, ]
  cat(sprintf(template, current$name, current$data, current$text))
}
```
